# Maximum Likelihood {#sec-max-lik}

## Lecture material

Please download and print the lecture materials from [Bblearn](https://bblearn.nau.edu/){target="_blank"}. After lectures, the recordings will appear in the Bblearn Collaborate Ultra section.

## Generate some data {#sec-data}

First, let's generate some data for the case of a simple linear regression.

```{r}
### PARAMS
beta0 = 1.5
beta1 = 0.5
sigma = 0.4
n = 30

### GENERATE DATA
set.seed(5)
x = runif(n, -1.5, 1.5)
exp_y = beta0 + beta1*x
y = exp_y + rnorm(n, mean=0, sd=sigma)

# Create a data frame 
my_df = data.frame(y = y, x = x)

plot(my_df$y ~ my_df$x, pch = 19,
     xlab = "x", ylab = "y")
```

## Calculate a likelihood

Remember that, for simple linear regression, the likelihood of a single data point is as follows:
$$y_i \sim N(\beta_0 + \beta_1 x_i, \sigma^2)$$
$$P(y_i | X_i B, \sigma^2) = \frac{1}{\sqrt{2 \pi \sigma^2}} \text{e}^{-\frac{1}{2}\frac{(y_i - X_i B)^2}{\sigma^2}}$$

Then, the full likelihood of the data set $Y$ is computed as:

$$ P(Y | X B, \sigma^2) = \prod^n_{i=1} P(y_i | X_i B, \sigma^2)$$
Or, on the natural logarithmic scale:
$$ ln\left(P(Y | X B, \sigma^2)\right) = \sum^n_{i=1} ln\left(P(y_i | X_i B, \sigma^2)\right)$$

```{r}
# How to calculate the likelihood of a single data point:
dnorm(y[1], 
      mean = beta0 + beta1*x[1],
      sd = sigma,
      log = FALSE)

# Calculate the full likelihood of the data, using the product
## Vectorized:
LH_notlog= 
prod(
    dnorm(y, 
          mean = beta0 + beta1*x,
          sd = sigma,
          log = FALSE)
)

# Log-likelihood, vectorized
LH_log = 
sum(
    dnorm(y, 
          mean = beta0 + beta1*x,
          sd = sigma,
          log = TRUE)
)

# Make sure the output makes sense
LH_notlog
# Indeed the log-likelihood is the log of the 
# likelihood on the raw probability scale.
LH_log; log(LH_notlog)
```

## Using `optim()` to minimize the negative log-likelihood

As we discussed in lecture, it is more computationally convenient to minimize functions, rather than to maximize. Therefore, to conduct linear regression analysis with maximum likelihood methods, we will find the values of $\hat{B}$ that minimize the negative log-likelihood of the data: $-ln\left(P(Y | X B, \sigma^2)\right)$.

### Function to calculate the negative log-likelihood

First, we need to construct a function that calculates the negative log-likelihood and that specifies the parameters of the model that eventually need to be estimated by the `optim()` function.

```{r}
### NEG LOG-LIK MINIMIZATION
neg_log_lik = function(p, data_df){
    beta0=p[1]
    beta1=p[2]
    sigma=p[3]
    
    mu = beta0 + beta1*data_df$x
    
    nll = -sum(dnorm(data_df$y, mean=mu, sd=sigma, log = TRUE))
    return(nll)
}
```
Here you can see the inputs to the function: `p` is a vector of parameters to be estimated (i.e., optimized), and `data_df` is a `data.frame` that holds the values of outcome variable $y$ and associated input variables, in this case, just one $x$. 

Then, we can use the `optim()` function, which implements a gradient descent algorithm to estimate the values of the parameters that minimize the provided function, `neg_log_lik()`. We will learn more about gradient descent later, because this is a very important method used widely across machine learning and neural networks. 

```{r}
m_nll = 
    optim(
        par = c(0.1,0,0.1),
        fn = neg_log_lik,
        data_df = my_df,
        method = "L-BFGS-B",
        lower=c(-5,-5,0.001),
        upper=c(5,5,5),
        hessian = TRUE
    )
par_tab_nll = rbind(m_nll$par)
colnames(par_tab_nll) = c("int", "slope", "sigma")
par_tab_nll
```
Note the `optim()` options. `par` specifies the initial guesses of the three parameters, whereas `lower` and `upper` specify the bounds across which to search for the best values of the parameters. With `hessian = TRUE` we are asking the function to output an estimate of the Hessian matrix of the function, which helps us to estimate the standard errors of the parameter estimates (see [Footnotes @sec-hessian]). The `method` specifies the algorithm used to minimize the function, which in this case is a modified quasi-Newton method, `L-BFGS-B`, which is a type of gradient descent algorithm, to be discussed later. 

We can see that the function outputs three point-estimates, which are $\hat{B}$ (i.e., slope and intercept), as well as the residual standard deviation, $\hat{\sigma}$.

### Compare `optim()` results to the OLS output
```{r}
### COMPARE TO LM()
m_ols = lm(y ~ 1 + x, data = my_df)
m_ols_summary = summary(m_ols)
m_ols_summary # Notice p-value

par_tab_ols = c(coef(m_ols), m_ols_summary$sigma)
names(par_tab_ols) = c("int", "slope", "sigma")
par_tab_ols; par_tab_nll

plot(my_df$y ~ my_df$x, pch = 19,
     xlab = "x", ylab = "y")
# Line from OLS
abline(coef = coef(m_ols), col = "black", lwd = 2)
# Line from MaxLikelihood
abline(coef = m_nll$par[1:2], col = "red", lty = 2, lwd = 2)
```

As we learned in lecture, the estimates of $\hat{B}$ from least squares and maximum likelihood are equivalent. And indeed, we see the same estimates produced from `lm()` and `optim()`. Also if you look at `Residual standard error` in the `lm()` output, you see the equivalent estimate for $\hat{\sigma}$ compared to the `optim()` output. 

## Hypothesis-testing for maximum likelihood

As explained in lecture, we will use the likelihood ratio test to test:

$$H_0: \beta_i = 0$$
$$H_A: \beta_i \ne 0$$
In the least squares framework, we used a $t$-test. But, for maximum likelihood, we are going to base our test on the *likelihood* of a model that does or does not include the slope, similar to the $F$-test we learned before. 

For the case of simple linear regression, we're testing whether there is a significant difference between these models:
$$H_0: y_i = \beta_0 + \epsilon_i$$
$$H_A: y_i = \beta_0 + \beta_1 x_i + \epsilon_i$$
Notice that in $H_0$, the slope $\beta_1$ is assumed to be zero.  

### Likelihood ratio and the $\chi^2$ test

Our goal is to understand if the likelihood of the null model, $P(Y | \beta_0, \sigma^2)$, is sufficiently low compared to the likelihood of the full model, $P(Y | \beta_0, \beta_1, x, \sigma^2)$, that we can reliably reject the null hypothesis.

We therefore construct a ratio of the likelihoods of the full and null model, very similar to the $F$-test framework. The log-likelihood ratio ($LHR$) becomes our test statistic:
$$LHR_{\text{test}} = -2 ln \left(\frac{LH_{\text{null}}}{LH_{\text{full}}} \right)$$

Then, folks smarter than I have done the math to prove that this test statistic is equivalent to a $\chi^2$ test statistic, such that:

$$LHR_{\text{test}} \sim  \chi^2_k$$

where $\chi^2_k$ is a $\chi^2$ probability distribution with $k$ degrees of freedom. $k$ is equal to $p_{\text{full}} - p_{\text{null}}$, where $p$ is the number of model coefficients. In the case of simple linear regression, where we are removing just one model coefficient from the full model (i.e., set slope equal to zero), then $k = 2-1 = 1$.

Finally, we can determine $P(\chi^2 > LHR_{\text{test}})$, which gives us our $p$-value. This statistical test is known as the "likelihood ratio test," and it is equivalently referred to as the "$\chi^2$" test, which we'll see in the code below. 

### Manual calculation of the likelihood ratio test

To begin, we need to use maximum likelihood to estimate the likelihood of the "null" model. We need to adjust our function that will be used by `optim()` to only include two parameters: the intercept, and the residual standard deviation. 

```{r}
# Need a null model:
nll_null = function(p, data_df){
    beta0=p[1]
    sigma=p[2]
    
    mu = beta0
    
    nll = -sum(dnorm(data_df$y, mean=mu, sd=sigma, log = TRUE))
    return(nll)
}

m_nll_null = 
    optim(
        par = c(0.1,0.1),
        fn = nll_null,
        data_df = my_df,
        method = "L-BFGS-B",
        lower=c(-5,0.001),
        upper=c(5,5)
    )
par_tab_nll_null = rbind(m_nll_null$par)
colnames(par_tab_nll_null) = c("int", "sigma")
par_tab_nll_null

plot(my_df$y ~ my_df$x, pch = 19,
     xlab = "x", ylab = "y")
abline(coef = m_nll$par[1:2], col = "red", lty = 1)
abline(h = m_nll_null$par[1], col = "red", lty = 2)
```

The flat dashed line represents the null (intercept-only) model. Now, we calculate the likelihood ratio test statistic, and compare to the $\chi^2$ probability distribution to determine our $p$-value of the test. Note that within the `optim()` function's output list, there is a numeric object called `value`. This `value` is the negative log-likelihood of the model with the estimated coefficients. We can use this to calculate our test statistic.

```{r}
# use exp() to convert the negative log likelihood to 
# raw probability scale
log_lh_full = -m_nll$value
log_lh_null = -m_nll_null$value

lh_full = exp(log_lh_full)
lh_null = exp(log_lh_null)

# Now calculate LHR
lhr = -2 * log(lh_null / lh_full)
lhr

# Of course, using rules of natural logs, this is equivalent:
-2 * (log_lh_null - log_lh_full)
```

Now that we have our value of $LHR_{\text{test}}$, we use the $\chi^2$-distribution to find $P(\chi^2 > LHR_{\text{test}})$, which is the $p$-value of the test.

```{r}
# How many parameters being "removed" (i.e., set to zero) in test:
df_chi = 2 - 1

# Prob null is true
p_val = 1 - pchisq(lhr, df = df_chi)
p_val
```
Based on this low $p$-value, we would say there is sufficient evidence to reject the null hypothesis and that the slope $\beta_1$ is significantly different than zero. 

We can compare this outcome to a built-in `R` function called `drop1()`. 

```{r}
drop1(m_ols, test = "Chisq")
```

In the function, we specified `Chisq` test, which implements the $\chi^2$ test using the likelihood ratio. What we see in this summary output is `Pr(>Chi)` which is equivalent to our manually computed value of $P(\chi^2 > LHR_{\text{test}})$. This output from `drop1()` does not provide a whole lot of detail, but if you look at the `help()`, it says that if you specify `test = "Chisq"`, it conducts a likelihood-ratio test. It doesn't specifically output the likelihood ratio, but we can see the $p$-value is equivalent to our manual calculation above. 

## Footnotes 

### Hessian matrix {#sec-hessian}

The `optim()` function provides point estimates for the maximum likelihood-derived model coefficients. Just like in least squares regression, however, we want to quantify the uncertainty in these estimates. We therefore want the standard error in the model coefficient estimates. 

In the case of least squares, we showed how we can calculate a variance-covariance matrix for the model coefficients, and then the square-root of the diagonal of this matrix equals the standard error. For maximum likelihood we can estimate this same variance-covariance matrix, but it comes from a different matrix called the Hessian. We do not need to go into detail, but the Hessian is the matrix of second derivatives of the likelihood with respect to the parameters (I will not ask you to recall this information). Then the variance-covariance matrix of the estimated model coefficients is calculated as the inverse of the Hessian matrix that corresponds to the negative log-likelihood. If the Hessian matrix of the negative log-likelihood is $H$, then
$$SE(\hat{\beta_i}) = \sqrt{\text{diag}\left( H^{-1}\right)_i}$$
I understand that's complicated, but it's easy enough to extract these values computationally from `optim()` output, assuming you use the option `hessian = TRUE`. 
```{r}
# Extract the Hessian from the optim() output
hessian = m_nll$hessian
# Calculate the var-cov matrix from the inverse Hessian
# Remember solve(X) gives X^-1
params_varcov = solve(hessian)
# Then extract the diagonal and take the square root
# This gives a vector of SE(\param_i)
se_params = sqrt(diag(params_varcov))
params_tab = cbind(m_nll$par, se_params)
colnames(params_tab) = c("Estimate", "Std. Error")
rownames(params_tab) = c("Intercept", "slope", "sigma")
params_tab

# Same as OLS? 
m_ols_summary$coefficients[c(1:2), 1:2] # Pretty close!
```
We can see that the standard errors for the maximum likelihood estimators are the same as the OLS estimators. 

### `optim()` using least squares {#sec-least-sq}

Remember that `optim()` is not specific to maximum likelihood, but rather it implements one of several optional minimization algorithms. Therefore, we can use it to minimize any quantity. To emphasize this point, remember that in least squares regression, we are finding the values of the model coefficients $\hat{B}$ that minimize the sum of squared errors, $\sum_i^n \epsilon_i^2 = \epsilon^T\epsilon$. Let's minimize this quantity using `optim()`. 

```{r}
### LEAST SQUARES MINIMIZATION

# We need a function to calculate the sum of squared errors:
least_sq = function(p, data_df){
    beta0=p[1]
    beta1=p[2]
    y = data_df$y
    n = length(y)
    
    expected_y = beta0 + beta1*data_df$x
    sse = 0
    for(i in 1:n){
        epsilon_i = y[i] - expected_y[i]
        sse = sse + (epsilon_i)^2
    }
    
    return(sse)
}

### OPTIMIZE LEAST SQUARES
fit_least_sq = 
    optim(
        par = c(0,0),
        fn = least_sq,
        data_df = my_df,
        method = "L-BFGS-B",
        lower=c(-5,-5),
        upper=c(5,5),
        hessian = TRUE
    )
# Create a table of estimates:
par_tab_least_sq = rbind(fit_least_sq$par)
colnames(par_tab_least_sq) = c("int", "slope")
par_tab_least_sq

# Compare to original OLS estimates:
coef(m_ols)
```

You could also use the Hessian output to calculate the standard errors of the model coefficients, but I will leave that up to you. 