---
title: "Problem Set 0"
subtitle: "INF 511"
author: "blank"
fig-align: center
fig-cap-location: bottom
format: 
    pdf: 
        documentclass: article
        geometry: 
          - top=1in
          - left=0.75in
          - bottom=1in
          - right=0.75in
editor: source
---

This is just a warm-up assignment for R and Quarto documents.

**You must turn this is as a rendered PDF to earn any credit for the assignment. See Question 4.**

# Q0: YAML

Change the `author` input of the YAML to be your name.

# Q1: Objects, arrays, loops

In the following code chunk, there is a skeleton of a `for` loop. In each element of the `store_array`, assign the squared value of the index. For instance, element two of the array would be assigned two-squared. Then extract the 5th element of the array.

```{r}
# Create a storage array (i.e., vector)
store_array = vector(mode = "numeric",
                     length = 10)

# You can see R fills the memory with zeros
store_array

# Assign an object on the fly
array_len = length(store_array)

# Indexing starts at position 1 in R
for(i in 1:array_len){
    
    # ENTER YOUR CODE HERE
}

# Extract the 5th element of the array:

```

# Q2: Plotting

Below, I create a data frame object with two columns. Create a scatter plot of the two columns. Use custom character strings to assign the x- and y-axis labels. Change the plotting character to something custom. Make the data points on the plot a different color from black.

```{r}
#| fig-width: 5
#| fig-height: 4.5

my_df = data.frame(
    y_var = c(1, 4, 3, 2, 8, 7, 1.5, 10, 9, 11),
    x_var = c(11:20)
)

# Use the plot() function
```

# Q3: Practice OLS with `lm()`

Use the same data frame as in the previous question. For the linear model, use the formula in the form `y_var ~ x_var`, and store the output of the `lm()` function into an object. Then, call the `summary()` function on that `lm` object, displaying the output here. Finally, indicate what are the estimates of the slope and intercept, which are displayed in the summary output.

```{r}
# Insert your code here
```

# Q4: Rendering

Render this document as a PDF file. Upload the rendered PDF onto BBLearn.
