---
title: "Problem Set 2"
subtitle: "INF 511"
author: "blank"
fig-align: center
fig-cap-location: bottom
number-sections: true
format: 
    pdf: 
        documentclass: article
        geometry: 
          - top=1in
          - left=0.75in
          - bottom=1in
          - right=0.75in
editor: source
---

# Manual OLS analysis to estimate $\hat{B}$

In this problem set, you will conduct OLS analysis on a data set that you will simulate. 

## Use the following parameters to generate the X covariate matrix.

First, you will need to generate the matrices and vectors that are needed to generate the data. Remember the $X$ matrix is $n$ x $p$, where $n$ is the number of data observations, and $p$ is the number of parameters. Here, $p=2$, for the intercept and one slope (i.e., we are only dealing with one covariate, $x$).

```{r}
n = 50
p = 2

x0 = rep(1, times = n)

# Randomly draw from a probability distribution to generate 
# n values for x1.
# x1 = ?

# Create the matrix, xmat, using the x0 and x1 column vectors

```

## Create an array of residual error values

Remember that $\epsilon_i \sim N(0, \sigma^2)$. Draw $\epsilon_i$ values randomly from a normal distribution.

```{r}
# Assign a value for sigma, the residual standard deviation
# sigma = ?

# Assign values for epsilon, drawing randomly from a normal distribution
# epsilon = ?
```

## Calculate the observed values of $Y$.

```{r}
# Use the following values of intercept and slope
# betas[1]: intercept
# betas[2]: slope
betas = c(1.50, 1.75)

# Calculate the values of y, using xmat, betas, and epsilon
# y = ?
```

## Calculate $\hat{B}$

Now we have the data observations. Using the example code already provided, calculate the coefficients that we estimate from the data using OLS, stored in matrix, $\hat{B}$. Use the `solve()` function.

```{r}
# Enter your code here
```

# OLS analysis using `lm()`

Use the `lm()` function to estimate the model coefficients. Store the estimated coefficients in an array.

```{r}
# Enter your code here
```

# Visualize and compare the analyses

Plot the observed data in a scatter plot using `plot()`. As in the code already provided, plot three lines: (1) linear relationship with true values of coefficients, $B$; (2) linear relationship with coefficients estimated from manual OLS analysis; and (3) linear relationship with coefficients estimated from `lm()`. Each line should be a different color and a different line type (option `lty` in the `abline()` function.) Finally, create chunk options (i.e., using the `#|` syntax) to specify the plot's height and width.

```{r}
# Enter your code here
```

# Rendering

Render this document as a `.PDF` file. Upload the rendered `.PDF` and the original `.QMD` onto BBLearn.
